Select-based random access \\ to variable-byte encodings

### 2020-04-24 gradu/gradu.tex
Updated introduction and abstract

Enormous datasets are a common occurence today and compressing them is often beneficial. Variable-byte encoding is a method for compressing integers of different byte lenghts. It removes the zero bytes 
and adds an additional \textit{continuation bit} to each byte to denote whether the compressed integer continues to the next byte or not. Two bit array data structures, $rank$ and $select$  are built on top of the array of 
continuation bits. These data structures enable a fast direct access to the elements in the compressed data. The $rank$ implementation performs better on data containing mostly small integers. The $select$ 
implementation benefits on larger integers. An experiment on fetching a subarray is explained and implemented. The $select$ implementation has significant advantages on subarray fetching due to how the data 
is compressed.

>>>>

Enormous datasets are a common occurence today and compressing them is often beneficial. Fast direct access to any element in the compressed data is a requirement in the field of compressed 
data structures, which is not possible with traditional compression methods.

Variable-byte encoding is a method for compressing integers of different byte lengths. It removes unused leading bytes and adds an additional \textit{continuation bit} to each byte to 
denote whether the compressed integer continues to the next byte or not. An existing solution using \textsc{rank} data structure performs well in this given task. This thesis introduces an 
alternative solution using \textsc{select} data structure and compares the two implementations. An experimentation is also done on retrieving a subarray from the compressed data structure.

The $rank$ implementation performs better on data containing mostly small integers. The $select$ implementation benefits on larger integers. The $select$ implementation has significant 
advantages on subarray fetching due to how the data is compressed.

### 2020-04-24 gradu/msc_english_contents.tex
more references
Updated introduction and abstract

Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuous bit} is added to the data to denote whether the compressed integer
is continued to the next byte or not. A light-weight data structure is built over the continuous bit array. The fact that the compressed data blocks are of the same size is used
with the data structure to allow direct access to the compressed variable-byte data.

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17, Sha16}.

@@

Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
very convenient and fast. A solution for subarray fetching and comparison is proposed in this thesis.

@@

\section{VB encoding}
The name \textit{variable-byte} is reflected in how the integers are stored. An array stores all the elements with the same bit length. This is very inefficient if a lot of the values are very small but the 
array also has to support very large integers. VB encoding attempts to eliminate the unneeded leading zeros. The bits in an integer are split into blocks of length $b$ and empty blocks from the beginning are 
discarded. Because the lengths of the integers may now be different, the data cannot be stored as it is. Instead a \textit{continuation bit} is added in front of each block to form a $chunk$. This bit is set 
to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer continues in the next chunk.

@@

\node[X, below=5ex of b1] (a2) {$A_{1,2}$};
\node [left=1ex of a2]{\textbf{$A_1$ =}};
\node[X] {$A_{3,2}$};
\node[X] {$A_{4,2}$};
\node[X] {...};

@@

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ now has the index of the element's next block in array $A_2$.
This process is repeated until the i-th bit of the bit array of the current level is set. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
with block length of 8.

@@

 - write more introduction

 - write more abstract

@@

 - Chapter 3 start: Direct access to any element.... explain why

 - 3.1 explain more what the bit array is

 - 3.1 faste => fast

@@

>>>>

Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuation bit} is added to the data to denote whether the compressed integer
is continued to the next byte or not. Two different light-weight data structures are built over the continuation bit array. $Rank_1$($i$) returns the number of 1 bits in the bit array between 
indexes 0 and $i$-1. $Select_1$($i$) returns the index of the i-th 1 bit in the array. The fact that the compressed data blocks are of the same size is used
with the data structures to allow direct access to the compressed variable-byte data.

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17, Sha16}.

@@

Decompressing a subarray is experimented later in this thesis. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a 
subarray decompression very convenient and fast. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared. $Select$ based method performs better due to its 
single $select$ call requirement, while the $rank$ call amount depends on the size of the largest element in the subarray.

@@

\section{VB encoding}
The name \textit{variable-byte} is reflected in how the integers are stored. When stored in an array, each element requires the same amount of space. This is very inefficient if a lot of the values are 
very small but the array also has to support very large integers. VB encoding attempts to eliminate the unneeded leading zeros. The bits in an integer are split into blocks of length $b$ and empty blocks 
from the beginning are discarded. Because the lengths of the integers may now be different, the data cannot be stored as it is. Instead a \textit{continuation bit} is added in front of each block to form 
a $chunk$. This bit is set to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer 
continues in the next chunk.

@@

\node[X, below=5ex of b1] (a2) {$A_{1,2}$};
\node [left=1ex of a2]{\textbf{$A_2$ =}};
\node[X] {$A_{3,2}$};
\node[X] {$A_{4,2}$};
\node[X] {...};

@@

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ is the index of desired element's next block in array $A_2$. This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current 
block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks.

A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

As an example, the fourth element from the data structure shown in Figure~\ref{bris_ds} is decoded. The first block of data is fetched by directly accessing $A_1$[3]. Then the 
continuation bit $B_1$[3] is checked. It is not set, which indicates that the element continues to the next block. The index of the next block is obtained via a \textsc{Rank}($B_1$, 3) call. There are two zero 
bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is set to 1 so the decompression is completed.

@@

@@

@@

 - explain rank/select in introduction

 - 1.1 research questions
